How are galaxies assigned to halos? \\ Searching for assembly bias in the SDSS galaxy clustering

### 2017-03-06 text/ms.tex
the most up-to-date version of the text

\newcommand{\acen}{\mathcal{A}_{\rm cen}}
\newcommand{\asat}{\mathcal{A}_{\rm sat}}
%\newcommand{\comment}[1]{{\bf \textcolor{red}{ #1}}}
\newcommand{\todo}[1]{{\em \textcolor{red}{ #1}}}

%% preprint2 produces a double-column, single-spaced document:

%% \documentclass[preprint2]{aastex}

%% Sometimes a paper's abstract is too long to fit on the
%% title page in preprint2 mode. When that is the case,
%% use the longabstract style option.

%% \documentclass[preprint2,longabstract]{aastex}

%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
%% If you are submitting to a journal that translates manuscripts
%% into SGML, you need to follow certain guidelines when preparing
%% your macros. See the AASTeX v5.x Author Guide
%% for information.

%% You can insert a short comment on the title page using the command below.

%\slugcomment{Not to appear in Nonlearned J., 45.}

%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
%% The left head contains a list of authors,
%% usually a maximum of three (otherwise use et al.).  The right
%% head is a modified title of up to roughly 44 characters.
%% Running heads will not print in the manuscript style.

%\shorttitle{Collapsed Cores in Globular Clusters}
%\shortauthors{Djorgovski et al.}

%% This is the end of the preamble.  Indicate the beginning of the
%% paper itself with \begin{document}.

@@

%% LaTeX will automatically break titles if they run longer than
%% one line. However, you may use \\ to force a line break if
%% you desire.
\title{How are galaxies assigned to halos? \\ Searching for assembly bias in the SDSS galaxy clustering}

%%\title{Collapsed Cores in Globular Clusters, \\
%%    Gauge-Boson Couplings, and AAS\TeX\ Examples}

%% Use \author, \affil, and the \and command to format
%% author and affiliation information.
%% Note that \email has replaced the old \authoremail command
%% from AASTeX v4.0. You can use \email to mark an email address
%% anywhere in the paper, not just in the front matter.
%% As in the title, use \\ to force line breaks.

\author{Mohammadjavad Vakili\altaffilmark{1}}

@@

Clustering of dark matter halos has been shown to depend on halo properties beyond mass such as halo concentration, a phenomenon referred to as halo assembly bias. Standard halo occupation modeling (HOD) in large scale structure studies assumes that halo mass alone is sufficient in characterizing the connection between galaxies and halos. Modeling of galaxy clustering can face systematic effects if the number or properties of galaxies are correlated with other halo properties. Using the Small MultiDark-Planck high resolution $N$-body simulation and the measurements of the projected two-point correlation function and the number density of Sloan Digital Sky Survey (SDSS) DR7 main galaxy sample, we investigate the extent to which the dependence of halo occupation on halo concentration can be constrained, and to what extent allowing for this dependence can improve our modeling of galaxy clustering.

@@

We show that in comparison with abundance-matching mock catalogs, our findings suggest qualitatively similar but modest levels of assembly bias that is only present in the central occupation. Furthermore, by performing model comparison based on information criteria, we find that in most cases, the standard mass-only HOD model is still favored by the observations.

@@

The complexity of structure formation however, is not sufficiently modeled under the standard HOD framework. Numerous $N$-body simulations that examine the clustering of 
dark matter halos have demonstrated that halo clustering is correlated with the formation 
history of halos. That is, at a fixed halo mass, the halo bias is correlated 
with properties of halos beyond mass, such as the concentration, formation time, or etc. 
This phenomenon is known as halo assembly bias (see \citealt{sheth2004,gao2005, harker2006, weschler2006, gao2007,croton2007,wang2007,angulo2008,dalal2008,li2008,sunayama2016}). 
There is support for halo assembly bias in observations of SDSS redMaPPer galaxy clusters (\citealt{miyatake2016}).

Furthermore, the properties of galaxies within the halo may also depend on the halo's formation history. Then we may expect the occupation statistics of galaxies to be tied to the halo properties beyond mass such as the concentration of halos. There have been many attempts in the literature at examining the dependence of galaxy properties on environment of the halos. But the results are mixed, and there is very little consensus. \citet{tinker_void_2009} show that the properties of the galaxies that reside in voids can be explained by the halo mass in which they live, and their properties are independent of their large scale environment of the halos. \citet{tinker_density_hod} proposes an extension of the standard HOD model $P(N|M)$ such that the number of galaxies residing in a halo not only depends on the mass of the halo, but also on the large scale density contrast $P(N|M,\delta)$. Based on modeling the clustering and void statistics of the SDSS galaxies, \citet{edHOD-tinker} shows that the dependence of the expected number of central galaxies on large scale density is not very strong.    
By randomly shuffling the galaxies among host halos of similar mass in the Millenium simulation, \citet{croton2007} shows that assembly bias significantly impacts the galaxy two-point correlation functions. They also show that the effect is different for the faint and the bright samples. By applying a group finder to SDSS DR7 galaxies, \citet{tinker2011} shows that the quenched fraction of galaxies is independent of the large-scale environment.

Based on a group catalog of SDSS galaxies, \citet{blanton_group_2007} find that shuffling of galaxy colors among groups of similar mass has very little effect (only a slight change in small scales: less than $\sim 300 \rm{kpc}$) on the clustering of galaxies. Based on an alternative group catalog, \citet{wang2013} shows that clustering of the central galaxies does not depend only on the halo mass, but also on additional parameters such as the specific star formation rate.

@@

Age Matching models extend the Abundance Matching models to account for even higher orders of galaxy 
assembly bias. In these models, galaxy color is matched to halo formation time so that older (redder) galaxies preferentially reside in older halos (\citealt{hw2013}). \citet{hearin2014} demonstrates that such an approach can reproduce a range of color and luminosity-dependent statistics of SDSS galaxies such as the 2PCF and galaxy-galaxy lensing. Based on the projected 2PCF measurements of (\citealt{hw2013}) galaxy catalogs, \citet{arz2014} showed that although the 2PCF measurements of these catalogs can be fit with the standard HOD modeling, the inferred HOD parameters cannot recover the underlying true HOD of these catalogs. That is, even in the presence of assembly bias in a galaxy sample, one can fit the clustering of this sample with HOD, but that does not guarantee recovery of the true HOD parameters. \citet{edHOD-weinberg} demonstrated that even with a galaxy catalog that has a significant level of assembly bias (e.g. \citealt{hw2013}), it is possible to extract the matter power spectrum more accurately if an environment-dependent HOD ($P(N|M,\delta)$) is used.

@@

Furthermore, we discuss how taking assembly bias into account in a more complex HOD model can improve our modeling of galaxy clustering in certain brightness limits. Then, we make a qualitative comparison between the levels of assembly bias in the mock catalogs created from our constraints on the decorated HOD model, and the \citealt{hw2013} mock catalogs created using the abundance matching technique that uses maximum circular velocity $V_{\rm max}$ as subhalo property. Our comparison shows the levels of assembly bias seen in the predictions of both models follow the same trend. That is assembly bias is more prominent in lower luminosity-threshold samples and its impact on galaxy clustering is only significant on large scales.

@@

@@

\citet{arz2014} shows that in the mock catalogs that exhibit significant levels of assembly bias, using a simple mass-only occupation model can lead to considerable biases in inference of the galaxy--halo connection parameters. Although we cannot rule out moderate level of assembly bias in our findings, we do not find any considerable discrepancy between the two models in terms of estimating the underlying HOD parameters.

A few galaxy--halo connection methods have been proposed in the literature that give rise to assembly bias in the galaxy population. \citet{arz2014} demonstrates that the abundance matching techniques based on $V_{\rm max }$ (as introduced in \citealt{hw2013,reddick2013}) exhibit some levels of assembly bias. 
We aim to provide a comparison between the levels of assembly bias in these mock catalogs and the mock catalogs predicted from our constraints on the decorated HOD model for $L_{\star}$-type galaxies. In particular, we consider the abundance matching catalogs produced by \citet{hw2013}. These catalogs have been extensively studied for examining potential systematic effects of galaxy assembly bias on cosmological (\citealt{edHOD-weinberg}) and halo occupation (\citealt{arz2014}) parameter inferences.

@@

For $L_{\star}$ galaxies, the difference in $w_{p}$ between the randomized and the original catalogs of \citet{hw2013} is shown in Figure \ref{fig:randomized} with the red curves. As demonstrated in Figure \ref{fig:randomized} (and as previously noted by \citealt{arz2014,edHOD-weinberg}), the relative difference in $w_{p}$ is only significant in relatively large scales. This implies that in these catalogs, only the central occupation are affected by assembly bias. This is in agreement with our findings.

Furthermore, we investigate whether the levels of assembly bias predicted by our findings are in agreement with the abundance matching catalogs of \citet{hw2013}. First, we make random draws from the posterior probability distribution function over the parameters of the model with assembly bias. Then, we create mock catalogs with these random draws, and then we compute the difference in $w_{p}$ between the randomized catalogs and the original catalogs. The relative difference in $w_{p}$ predicted from our constraints are shown with blue curves in Figure \ref{fig:randomized}.

We note that our findings follow the same trend. That is, we see negligible difference in the small scale clustering and more considerable differences in $w_{p}$ on larger scales ($r_{p} > 1 Mpc \; h^{-1}$). Similar to findings of \citet{arz2014}, \citet{lehman2015}, and \citet{edHOD-weinberg}, we see that assembly bias becomes less significant in brighter galaxy samples. Furthermore, we notice that our mock catalogs favor more moderate levels of assembly bias as the fractional difference in $w_{p}$ is smaller in our mock catalogs.

@@

For the $M_{\rm r}<-20.5, -20, -19.5$ samples, at a fixed halo mass, we find positive correlation between the central population and halo concentration at fixed halo mass. For $M_{\rm r}<-19, -18.5$ we find no correlation, and for the faintest sample, we find negative correlation. Given the large scale halo assembly bias and the scale-dependent feature of assembly bias in the quasi-linear scales, these constraints lead to improvement in modeling the galaxy clustering. However, we do not find these improvements to be sufficient to lower the information criteria scores associated with the more complex model. The exceptions are the $M_{\rm r}<-20,-18$ luminosity-threshold samples for which we find the strongest constraints on the central assembly bias. For these two samples, the HOD model with assembly bias yields lower BIC score than the model without assembly bias.

Comparison of the levels of assembly bias in the catalogs constructed based on our results and the abundance matching catalogs presented in \citet{hw2013,arz2014} and \citet{edHOD-weinberg} demonstrates that our constraints predict more moderate levels of assembly bias which follows the same trend. That is, assembly bias mostly affects the large scale and the quasi-linear clustering and the small scale clustering remains unaltered. In addition, the effect of assembly bias vanishes in the brightest galaxy samples.

@@

We are grateful to David~W.~Hogg, Chang~Hoon~Hahn, Alex~I.~Malz, Andrew~Hearin, Andrew~Zentner, Chia-Hsun~Chuang, Michael~R.~Blanton, and Kilian~Walsh for discussions related to this work. This work was supported by the NSF grant AST-1517237. All the computations in this work were carried out in the New York University High Performance Computing Mercer facility. We thank Shenglong~Wang, the administrator of the NYU HPC center for his continuous and consistent support throughout the completion of this study.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% WP RANDOMIZED
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[p]~\\
\begin{center}
\includegraphics[width=\textwidth]{paper_wprandom.pdf}
 \caption{Demonstration of the relative difference in $w_{p}$ between randomized and non-randomized catalogs for different luminosity threshold samples: $M_{\mathrm r}<-20,-20.5,-21$. The errorbars are from the diagonal elements of the covariance matrix. The blue lines correspond to the random draws from the posterior probability (summarized in Table \ref{tab:constraints}) over the parameters of the HOD model with assembly bias. The red line corresponds to the subhalo abundance matching catalog (\citealt{hw2013,hearin2014}). Comparison between the level of assembly bias (encoded in the relative difference in $w_p$) shows that the our constraints favor \emph{more} \emph{moderate} levels of assembly bias than the levels seen in the abundance matching mock catalogs. Within both models, the small scale clustering remains unaltered after randomizing the catalogs, signaling the lack of correlation between the satellite population and the halo concentration at a fixed mass in the two models.}
\label{fig:randomized}
\end{center}
\end{figure*}

>>>>

\newcommand{\acen}{\mathcal{A}_{\rm cen}}
\newcommand{\asat}{\mathcal{A}_{\rm sat}}
\newcommand{\todo}[1]{{\em \textcolor{red}{ #1}}}

@@

\title{How are galaxies assigned to halos? \\ Searching for assembly bias in the SDSS galaxy clustering}

\author{Mohammadjavad Vakili\altaffilmark{1}, Chang~Hoon Hahn\altaffilmark{1}}

@@

Clustering of dark matter halos has been shown to depend on halo properties beyond mass such as halo concentration, a phenomenon referred to as halo assembly bias. Standard halo occupation modeling (HOD) in large scale structure studies assumes that halo mass alone is sufficient in characterizing the connection between galaxies and halos. Modeling of galaxy clustering can face systematic effects if the number of galaxies are correlated with other halo properties. Using the Small MultiDark-Planck high resolution $N$-body simulation and the measurements of the projected two-point correlation function and the number density of Sloan Digital Sky Survey (SDSS) DR7 main galaxy sample, we investigate the extent to which the dependence of halo occupation on halo concentration can be constrained, and to what extent allowing for this dependence can improve our modeling of galaxy clustering.

@@

We show that in comparison with abundance-matching mock catalogs, our findings suggest qualitatively similar but modest levels of the impact of halo assembly bias on galaxy clustering. That is, the effect is only present in the central occupation and becomes less significant in brighter galaxy samples.
Furthermore, by performing model comparison based on information criteria, we find that in most cases, the standard mass-only HOD model is still favored by the observations.

@@

The complexity of structure formation however, is not sufficiently modeled under the standard HOD framework. Numerous $N$-body simulations that examine the clustering of 
dark matter halos have demonstrated that halo clustering is correlated with the formation 
history of halos. That is, at a fixed halo mass, the halo bias is correlated 
with properties of halos beyond mass, such as the concentration, formation time, or etc. 
This phenomenon is known as halo assembly bias (see \citealt{sheth2004,gao2005, harker2006, weschler2006, gao2007,croton2007,wang2007,angulo2008,dalal2008,li2008,sunayama2016}). It has been claimed that there is support for halo assembly bias in observations of SDSS redMaPPer galaxy clusters (\citealt{miyatake2016}).

Furthermore, the halo occupation may also depend on the formation history of halos. Then we may expect the spatial statistics of galaxies to be tied to the halo properties beyond mass such as the concentration of halos. There have been many attempts in the literature at examining the dependence of halo occupation on environment of the halos. But the results are mixed, and there is very little consensus. \citet{tinker_void_2009} show that the properties of the galaxies that reside in voids can be explained by the halo mass in which they live, and their properties are independent of their large scale environment of the halos. \citet{tinker_density_hod} proposes an extension of the standard HOD model $P(N|M)$ such that the number of galaxies residing in a halo not only depends on the mass of the halo, but also on the large scale density contrast $P(N|M,\delta)$. Based on modeling the clustering and void statistics of the SDSS galaxies, \citet{edHOD-tinker} shows that the dependence of the expected number of central galaxies on large scale density is not very strong.    
By randomly shuffling the galaxies among host halos of similar mass in the Millenium simulation, \citet{croton2007} shows that assembly bias significantly impacts the galaxy two-point correlation functions. They also show that the effect is different for the faint and the bright samples.

%Based on a group catalog of SDSS galaxies, \citet{blanton_group_2007} find that shuffling of galaxy colors among groups of similar mass has very little effect (only a slight change in small scales: less than $\sim 300 \rm{kpc}$) on the clustering of galaxies. Based on an alternative group catalog, \citet{wang2013} shows that clustering of the central galaxies does not depend only on the halo mass, but also on additional parameters such as the specific star formation rate.
%\todo{I think this paragraph might be a bit misleading to the reader, I think the paper could do without it. %Your work doesn't incorporate galaxy properties, but only halo properties.}

@@

Based on the projected 2PCF measurements of (\citealt{hw2013}) galaxy catalogs, \citet{arz2014} showed that after fitting the 2PCF measurements of these catalogs with the standard \emph{mass-only} HOD modeling, the inferred HOD does not match the \emph{true} halo occupation of these catalogs. That is, in the presence of assembly bias in a galaxy sample, one can fit the clustering of this sample with the standard \emph{mass-only} HOD, but that does not guarantee recovery of the true HOD parameters.

@@

Furthermore, we discuss how taking assembly bias into account in a more complex HOD model can improve our modeling of galaxy clustering in certain brightness limits. Then, we make a qualitative comparison between the levels of the impact of assembly bias in our best-fit decorated HOD model on galaxy clustering, and the impact of assembly bias present in (\citealt{hw2013}) catalogs on galaxy clustering. Our comparison shows the levels of the impact of assembly bias on galaxy clustering seen in the predictions of both models follow the same trend. That is assembly bias is more prominent in lower luminosity-threshold samples and its impact on galaxy clustering is only significant on large scales (more than a few Mpc).

@@

Note that a full forward model of the data requires running the simulation at different redshifts, generation of light cones, accounting for the complex survey geometry and systematic errors such as fiber collisions. Using the $z=0$ output of the $\mathtt{SMDP}$ simulation in our forward model of the spatial distribution of galaxies is only an approximation. This approximation can be justified by the small redshift range of the SDSS DR7 main galaxy sample. As described in \citealt{zehavi2011}, using random catalogs with angular window function of the data in measurements of galaxy clustering accounts for the geometry of the data. As described in Section~\ref{sec:data}, the fiber collision correction method of \citealt{guo2012} is applied to the SDSS clustering measurements used in this study. Therefore we do not account for that effect in our forward model.

@@

\citet{arz2014} shows that in the mock catalogs that exhibit significant levels of assembly bias, using a simple mass-only occupation model can lead to considerable biases in inference of the galaxy--halo connection parameters. Although we cannot rule out moderate levels of assembly bias in our findings, we do not find any considerable discrepancy between the two models in terms of estimating the underlying HOD parameters.

A few galaxy--halo connection methods have been proposed in the literature that give rise to assembly bias in the galaxy population. \citet{arz2014} demonstrates that the abundance matching techniques based on $V_{\rm max }$ (as introduced in \citealt{hw2013,reddick2013}) exhibit some levels of assembly bias. 
We aim to provide a comparison between the impact of assembly bias on clustering in these mock catalogs and the mock catalogs predicted from our constraints on the decorated HOD model for $L_{\star}$-type galaxies. In particular, we consider the abundance matching catalogs produced by \citet{hw2013}. These catalogs have been extensively studied for examining potential systematic effects of galaxy assembly bias on cosmological (\citealt{edHOD-weinberg}) and halo occupation (\citealt{arz2014}) parameter inferences.

@@

For $L_{\star}$ galaxies, the difference in $w_{p}$ between the randomized and the original catalogs of \citet{hw2013} is shown in Figure \ref{fig:randomized} with the red curves. As demonstrated in Figure \ref{fig:randomized} (and as previously noted by \citealt{arz2014,edHOD-weinberg}), the relative difference in $w_{p}$ is only significant in relatively large scales ($r_{p} > 1 \mathrm{Mpc} \; h^{-1}$). This implies that in these catalogs, only the central occupation is affected by assembly bias. This is in agreement with our findings.

Furthermore, we investigate whether the impact of assembly bias on galaxy clustering predicted by our findings are in agreement with the abundance matching catalogs of \citet{hw2013}. First, we make random draws from the posterior probability distribution function over the parameters of the model with assembly bias. Then, we create mock catalogs with these random draws, and then we compute the difference in $w_{p}$ between the randomized catalogs and the original catalogs. The relative difference in $w_{p}$ predicted from our constraints are shown with blue curves in Figure \ref{fig:randomized}.

We note that our findings follow the same trend. That is, we see negligible difference in the small scale clustering and more considerable differences in $w_{p}$ on larger scales ($r_{p} > 1\; \mathrm{Mpc}h^{-1}$). Similar to findings of \citet{arz2014}, \citet{lehman2015}, and \citet{edHOD-weinberg}, we see that the impact of assembly bias on galaxy clustering becomes less significant in brighter galaxy samples. Furthermore, we notice that our mock catalogs favor more moderate changes in galaxy clustering as a result of assembly bias.

@@

For the $M_{\rm r}<-20.5, -20, -19.5$ samples, at a fixed halo mass, we find positive correlation between the central population and halo concentration at fixed halo mass. For $M_{\rm r}<-19, -18.5$ we find no correlation, and for the faintest sample, we find negative correlation. Given the large scale halo assembly bias and the scale-dependent feature of assembly bias in the quasi-linear scales, our constraints on the more flexible HOD model lead to improvement in modeling the galaxy clustering. However, we do not find these improvements to be sufficient to lower the information criteria scores associated with the more complex model. The exceptions are the $M_{\rm r}<-20,-18$ luminosity-threshold samples for which we find the strongest constraints on the central assembly bias. For these two samples, the HOD model with assembly bias yields lower BIC score than the model without assembly bias.

We compare the impact of assembly bias on galaxy clustering between the catalogs constructed from our results and the abundance matching catalogs presented in \citet{hw2013,arz2014}. We demonstrate that the effect of assembly bias on galaxy clustering predicted from our results is similar to (but more moderate than) the effects seen in the abundance matching catalogs of \citet{hw2013}. That is, assembly bias mostly affects the large scales and the quasi-linear clustering and the small scale clustering remains unaltered. In addition, the effect of assembly bias on galaxy clustering vanishes in the brightest galaxy samples.

@@

We are grateful to David~W.~Hogg, Alex~I.~Malz, Andrew~Hearin, Andrew~Zentner, Chia-Hsun~Chuang, Michael~R.~Blanton, and Kilian~Walsh for discussions related to this work. This work was supported by the NSF grant AST-1517237. All the computations in this work were carried out in the New York University High Performance Computing Mercer facility. We thank Shenglong~Wang, the administrator of the NYU HPC center for his continuous and consistent support throughout the completion of this study.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% WP RANDOMIZED
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[p]~\\
\begin{center}
\includegraphics[width=\textwidth]{paper_wprandom.pdf}
 \caption{Demonstration of the relative difference in $w_{p}$ between randomized and non-randomized catalogs for different luminosity threshold samples: $M_{\mathrm r}<-20,-20.5,-21$. The errorbars are from the diagonal elements of the covariance matrix. The blue lines correspond to the random draws from the posterior probability (summarized in Table \ref{tab:constraints}) over the parameters of the HOD model with assembly bias. The red line corresponds to the subhalo abundance matching catalog (\citealt{hw2013,hearin2014}). Our constraints favor \emph{more} \emph{moderate} levels of the impact of assembly bias on galaxy clustering than the levels seen in the abundance matching mock catalogs. Within both models, the small scale clustering remains unaltered after randomizing the catalogs, signaling the lack of correlation between the satellite occupation and the halo concentration at a fixed mass in the two models.}
\label{fig:randomized}
\end{center}
\end{figure*}
